---
title: "PS5 Markdown"
author: "Rahel Worku"
date: "2023-02-18"
output: html_document
---


### 1 Load and Check Data
```{r message=FALSE}
library(tidyverse)
gapminder <- read_delim("data/gapminder.csv.bz2")
dim(gapminder)
```
A sample of data:
```{r message=FALSE}
gapminder %>% 
  sample_n(5)
```

### 2 Descriptive statistics
1)
```{r}
iso2 <- distinct(gapminder, iso2)
iso3 <- distinct(gapminder, iso3)
name <- distinct(gapminder, name)
iso2
iso3
name

gapminder %>% 
  select(iso2, iso3, name) %>% 
  mutate(alliso2 = n_distinct(iso2),
         alliso3 = n_distinct(iso3),
         allname = n_distinct(name))
```

2a)
```{r}
names_per_iso2 <- gapminder %>%
  group_by(iso2) %>%
  summarize(num_names = n_distinct(name)) %>%
  arrange(desc(num_names))
names_per_iso2
  
duplicated_names_per_iso2 <- names_per_iso2[names_per_iso2$num_names > 1,]
duplicated_names_per_iso2

```
2b)
```{r}
iso3_per_name <- gapminder %>%
  group_by(name) %>%
  summarize(num_iso3 = n_distinct(iso3)) %>%
  arrange(desc(num_iso3))
iso3_per_name

duplicated_iso3_per_name <- iso3_per_name[iso3_per_name$num_iso3 > 1,]
duplicated_iso3_per_name
```
3)
```{r}
min_year <- min(gapminder$time, na.rm = TRUE)
max_year <- max(gapminder$time, na.rm = TRUE)
min_year
max_year
```

### 3 CO2 emissions

1)
```{r}
co2_missing <- gapminder %>% 
  group_by(time) %>% 
  summarize(n_missing = sum(is.na(co2))) %>% 
  arrange(desc(n_missing)) %>% 
  slice_head()

co2pc_missing <- gapminder %>% 
  group_by(time) %>% 
  summarize(n_missing = sum(is.na(co2_PC))) %>% 
  arrange(desc(n_missing)) %>% 
  slice_head()

co2_missing
co2pc_missing
```

2)
```{r message=FALSE, warning=FALSE}
countries <- c("United States of America", "China", "India", "Japan", "Malta")
country_data <- gapminder %>% 
  filter(name %in% countries) %>% 
  select(name, time, co2)
country_data

ggplot(country_data, aes(x=co2, y=time, col= factor(name))) +
  geom_line() +
  labs(x = "CO2 emissions", y = "Year", 
       col = "Country")

```
I can see from the plot that China has the most amount of emissions that are
steady increasing. The United states had the highest jump in co2 emissions 
though, but China overtook the US and continues to grow in their emissions.

3)
```{r message=FALSE, warning=FALSE}
countries <- c("United States of America", "China", "India", "Japan", "Malta")
country_data_pc <- gapminder %>% 
  filter(name %in% countries) %>% 
  select(name, time, co2_PC)
country_data_pc

ggplot(country_data_pc, aes(x=co2_PC, y=time, col= factor(name))) +
  geom_line() +
  labs(x = "CO2 emissions per capita", y = "Year", 
       col = "Country")
```
I can see that China and the United States contributed the most the co2
emissions per captia, with India as a close third to them.

4)
```{r}
gapminder %>%
  group_by(region) %>%
  summarise(mean_co2_pc = mean(co2_PC, na.rm = TRUE)) %>% 
  arrange(rank(desc(mean_co2_pc)))
```
The countries in the European and American regions are the countries with
the higher co2 emissions per captia

5)
```{r message=FALSE, warning=FALSE}
avg_co2_pc <- gapminder %>%
  filter(!is.na(co2_PC)) %>%
  filter(!is.na(region)) %>%
  group_by(time, region) %>%
  summarize(mean_co2_pc = mean(co2_PC))

avg_co2_pc <- avg_co2_pc %>%
  filter(time %in% c(1960, 2016))
avg_co2_pc

ggplot(avg_co2_pc, aes(x = region, y = mean_co2_pc, fill = factor(time))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Year", y = "Average CO2 Emissions per Capita", fill = "Region") +
  ggtitle("Average CO2 Emissions per Capita by Continent")
```

6) COMMENT
```{r message=FALSE, warning=FALSE}
co2_by_continent <- gapminder %>% 
  filter(time == 2019) %>% 
  group_by(region, time)
co2_by_continent

top_co2_by_continent <- co2_by_continent %>%
  arrange(desc(co2_PC)) %>% 
  group_by(region) %>%
  slice_head(n = 3)

bottom_co2_by_continent <- co2_by_continent %>% 
  arrange(co2_PC) %>% 
  group_by(region) %>%
  slice_head(n = 3)
top_co2_by_continent 
bottom_co2_by_continent

# I couldn't get the top and bottom three results even after trying the head
# function or by slicing.
```

### 4 GPD per capita

1)
```{r}
gapminder_1960 <- gapminder %>% 
  filter(time == "1960") %>% 
  group_by(name, totalPopulation, region) %>% 
  summarise(avg_life_expectancy = mean(lifeExpectancy, na.rm = TRUE),
            avg_gdp = mean(GDP_PC, na.rm = TRUE)) %>% 
  filter(!is.na(avg_life_expectancy)) %>% 
  filter(!is.na(avg_gdp)) %>% 
  filter(!is.na(name))
gapminder_1960

ggplot(gapminder_1960, aes(x = avg_gdp, y = avg_life_expectancy, size = totalPopulation, color = region)) +
  geom_point(alpha = 0.7) +
  scale_color_discrete(name = "Continent") +
  scale_size(range = c(1, 10), breaks = c(1e6, 1e7, 1e8, 1e9),
             labels = c("1M", "10M", "100M", "1B")) +
  labs(x = "GDP per capita (constant 2010 US$)",
       y = "Life expectancy (years)",
       title = "GDP per capita versus Life Expectancy by Country (1960)",
       subtitle = "Point size is proportional to population size") +
  theme_bw()
```



The scatter plot shows a positive relationship between GDP per capita and life 
expectancy, which is expected since higher GDP per capita is generally 
associated with better access to healthcare, nutrition, and living conditions. 

2)
```{r message=FALSE, warning=FALSE}
gapminder_2019 <- gapminder %>% 
  filter(time == "2019") %>% 
  group_by(name, totalPopulation, region) %>% 
  summarise(avg_life_expectancy = mean(lifeExpectancy, na.rm = TRUE),
            avg_gdp = mean(GDP_PC, na.rm = TRUE)) %>% 
  filter(!is.na(avg_life_expectancy)) %>% 
  filter(!is.na(avg_gdp)) %>% 
  filter(!is.na(name))
gapminder_1960

ggplot(gapminder_2019, aes(x = avg_gdp, y = avg_life_expectancy, size = totalPopulation, color = region)) +
  geom_point(alpha = 0.7) +
  scale_color_discrete(name = "Continent") +
  scale_size(range = c(1, 10), breaks = c(1e6, 1e7, 1e8, 1e9),
             labels = c("1M", "10M", "100M", "1B")) +
  labs(x = "GDP per capita (constant 2010 US$)",
       y = "Life expectancy (years)",
       title = "GDP per capita versus Life Expectancy by Country (2019)",
       subtitle = "Point size is proportional to population size") +
  theme_bw()
```


3)

Comparing the two plots, we can see a clear trend of global economic growth and improvement in life expectancy over the last 60 years. In 1960, there were clear differences between countries in terms of GDP per capita and life expectancy.
There is still a lot of variation between countries and regions, showing that
there's work to be done to achieve global economic and health equality.

4)
```{r message=FALSE, warning=FALSE}
gapminder %>%
  filter(time %in% c(1960, 2019)) %>%
  group_by(time, region) %>%
  summarize(avg_life_exp = mean(lifeExpectancy, na.rm = TRUE))
```
These results fit with the results of the scatter plots showing how the average
life expectancy in different regions have improved between the year 1960 till
2019.

5)
```{r}
LE_growth <- gapminder %>% 
  filter(time %in% c(1960, 2019)) %>% 
  group_by(region) %>% 
  mutate(LE_growth = last(lifeExpectancy) - first(lifeExpectancy))

continent_LE_growth <- LE_growth %>% 
  group_by(region) %>% 
  summarize(avg_LE_growth = mean(LE_growth))
continent_LE_growth
```
Countries in Africa had the highest growth in life expectancy. This makes sense
since healthcare has gotten more advanced since 1960 which has helped increase
the life expectancy in regions that were struggling. The more advanced countries
in Europe had a smaller raise in life expectancy since they've always had a
better life expectancy in general.

6)
```{r message=FALSE, warning=FALSE}
#gdp_data <- gapminder %>% 
#  filter(time %in% c(1960, 2019)) %>%
# group_by(name) %>% 
# select(name, GDP_PC, time)
#gdp_data

#ggplot(gdp_data, aes(x = GDP_PC ,fill = factor(name))) +
#  geom_histogram(alpha = 0.5, position = "identity", bins = 30) +
#  scale_fill_manual(values = c("#66c2a5", "#fc8d62")) +
# labs(title = "Distribution of GDP per capita in 1960 and 2019", x = "GDP per capita", y = "Frequency") +
#  theme_bw()

# I tried to get the the GDP_PC for the years 1960 and 2019 and I was going to
# compare their values in the histogram, but I couldn't fix the plot to overlay
# the data from both years. 
```

7)
```{r}
LE_rank_1960 <- gapminder %>% 
  filter(time == 1960,
         !is.na(name)) %>% 
  arrange(desc(-lifeExpectancy)) %>% 
  mutate(ranking = rank(-lifeExpectancy)) %>% 
  select(name, time, ranking) %>% 
  filter(name == "United States of America")
LE_rank_1960

LE_rank_2019 <- gapminder %>% 
  filter(time == 2019,
         !is.na(name)) %>% 
  arrange(desc(-lifeExpectancy)) %>% 
  mutate(ranking = rank(-lifeExpectancy)) %>% 
  select(name, time, ranking) %>% 
  filter(name == "United States of America")
LE_rank_2019
```

8)
```{r message=FALSE, warning=FALSE}
#gapminder %>%
 # filter(time %in% c(1960, 2019)) %>%
  #summarize(num_countries = n_distinct(name)) %>% 
  #mutate(rank_LE = rank(gapminder$lifeExpectancy)) %>%
  #mutate(relative_rank = rank_LE / num_countries) %>% 
  #filter(name == "United States of America") %>%
  #select(time, rank, num_countries, relative_rank)


#num_countries <- gapminder %>%
 # filter(!is.na(lifeExpectancy)) %>%
  #group_by(time) %>%
  #summarize(num_countries = n())
#num_countries

#relative_rank <- gapminder %>%
 # filter(!is.na(lifeExpectancy)) %>%
  #group_by(time) %>%
  #mutate(rank = dense_rank(-lifeExpectancy) / num_countries)
#relative_rank

# I tried different ways of getting the relative rate divided by the number of 
# countries in a certain year, but kept getting an error with my rank length.
# I was going to get the total countries in each year, and divide the US life
# expectancy rank by that number.
```

